---
title: "Case Study 3: Biomarkers for Predicting Prostate Cancer Recurrence"
author: "Emmanuel Fle Chea"
date: "`r Sys.Date()`"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
# Load necessary libraries
library(dplyr)       # Data manipulation
library(ggplot2)     # Plotting
library(glmnet)      # Penalized logistic regression (LASSO)
library(pROC)        # ROC curves, AUC, and DeLong test
library(knitr)       # Nice tables
library(caret)       # Cross-validation and data splitting
```

## Data Loading and Preprocessing

```{r, message=FALSE, warning=FALSE}
genes40 <- read.csv("C:/Users/emman/OneDrive/Desktop/MPH Data Science - UMN/Biostatistical Consulting/Case_Study3/casestudy3_40genes_data.csv",
                    stringsAsFactors = FALSE)

# Check the structure of the data
str(genes40)
```


```{r}
# Check data summary
colnames(genes40)
```


```{r}
# Loot at gleason
genes40$gleason
```

## Recode Variables and Define the Outcome

```{r}
# Recode 'gleason' as an ordered factor.
# Note: The resulting polynomial contrasts (e.g., gleason.L, gleason.Q, etc.) 
# reflect the inherent clinical ordering of Gleason scores.
genes40$gleason <- factor(
  genes40$gleason, 
  levels = c("3+3", "3+4", "4+3", "4+4", "4+5"),
  ordered = TRUE
)

# Create a binary outcome variable: recurrence within 5 years.
# Here, patients with biochem_failure == 1 and followup_time <= 5 are coded as 1 (Recurrence).
# Patients with followup times less than 5 years but with no event are considered "No Recurrence"
# because no biochemical failure was observed within the available follow-up period.
genes40 <- genes40 %>%
  mutate(
    recurrence_5yr = ifelse(biochem_failure == 1 & followup_time <= 5, 1, 0)
  )

# Convert the outcome to a factor with meaningful labels for caret
genes40$recurrence_5yr_factor <- factor(
  genes40$recurrence_5yr, 
  levels = c(0, 1),
  labels = c("NoRecurrence", "Recurrence")
)

# Verify the distribution of the outcome
table(genes40$recurrence_5yr_factor)
```


```{r}
# After recoding gleason, now check
genes40$gleason
```

## Data Splitting for Model Development and Evaluation

```{r}
# For reproducibility
set.seed(123)

# Split the data into training (70%) and test (30%) sets.
trainIndex <- createDataPartition(genes40$recurrence_5yr, p = 0.7, list = FALSE)
trainData <- genes40[trainIndex, ]
testData  <- genes40[-trainIndex, ]
```


## Exploratory Data Analysis (EDA)

```{r}
# Summary statistics for key clinical covariates on the full dataset
summary(genes40[, c("age", "preop_psa", "followup_time")])
table(genes40$gleason)
table(genes40$stage_t_n_m)

# Histograms for age and preoperative PSA
ggplot(genes40, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "grey", color = "black") +
  labs(title = "Histograms  for Distribution of Age", x = "Age (years)", y = "Count")

ggplot(genes40, aes(x = preop_psa)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black") +
  labs(title = "Histograms for Distribution of Preoperative PSA", x = "Preop PSA (ng/mL)", y = "Count")
```

## Table 1: Baseline Characteristics (Stratified by Recurrence Status)

```{r}
N <- nrow(genes40)
num_events <- sum(genes40$biochem_failure == 1, na.rm = TRUE)
median_followup <- median(genes40$followup_time, na.rm = TRUE)

mean_age <- mean(genes40$age, na.rm = TRUE)
sd_age   <- sd(genes40$age, na.rm = TRUE)

mean_psa <- mean(genes40$preop_psa, na.rm = TRUE)
sd_psa   <- sd(genes40$preop_psa, na.rm = TRUE)

gleason_tab <- table(genes40$gleason)
stage_tab   <- table(genes40$stage_t_n_m)

table1 <- data.frame(
  Variable = c(
    "Number of patients",
    "Number of events",
    "Median follow-up time (years)",
    "Age, mean (SD)",
    "Pre-operative PSA, mean (SD)",
    "Gleason score, n (%) 3+3",
    "Gleason score, n (%) 3+4",
    "Gleason score, n (%) 4+3",
    "Gleason score, n (%) 4+4",
    "Gleason score, n (%) 4+5",
    "Stage, n (%) Localized (0)",
    "Stage, n (%) Non-localized (1)"
  ),
  Value = c(
    N,
    num_events,
    round(median_followup, 3),
    sprintf("%.2f (%.2f)", mean_age, sd_age),
    sprintf("%.2f (%.2f)", mean_psa, sd_psa),
    paste0(gleason_tab["3+3"], " (", round(100 * gleason_tab["3+3"] / N, 1), "%)"),
    paste0(gleason_tab["3+4"], " (", round(100 * gleason_tab["3+4"] / N, 1), "%)"),
    paste0(gleason_tab["4+3"], " (", round(100 * gleason_tab["4+3"] / N, 1), "%)"),
    paste0(gleason_tab["4+4"], " (", round(100 * gleason_tab["4+4"] / N, 1), "%)"),
    paste0(gleason_tab["4+5"], " (", round(100 * gleason_tab["4+5"] / N, 1), "%)"),
    paste0(stage_tab["0"], " (", round(100 * stage_tab["0"] / N, 1), "%)"),
    paste0(stage_tab["1"], " (", round(100 * stage_tab["1"] / N, 1), "%)")
  ),
  stringsAsFactors = FALSE
)

kable(
  table1, 
  caption = "Table 1: Baseline Characteristics (Stratified by Recurrence Status)",
  align = c("l", "l")
)
```

## Model Development on the Training Set

### Model A: Clinical Covariates Only

```{r}
# Fit Model A (standard logistic regression) using training data
modelA_final <- glm(
  recurrence_5yr ~ preop_psa + age + gleason + stage_t_n_m,
  data = trainData, 
  family = "binomial"
)

# Display summary of Model A
summary(modelA_final)
```

## Model B: Clinical + Biomarkers with LASSO (Using 1-SE Rule)

```{r}
# Identify biomarker columns by excluding non-biomarker columns
non_biomarker_cols <- c(
  "case", "preop_psa", "gleason", "age", "stage_t_n_m",
  "followup_time", "biochem_failure", "recurrence_5yr", "recurrence_5yr_factor"
)
biomarker_names <- setdiff(names(trainData), non_biomarker_cols)
length(biomarker_names)  # Should be 40

# Create a formula for Model B
modelB_formula <- as.formula(
  paste("recurrence_5yr ~ preop_psa + age + gleason + stage_t_n_m +",
        paste(biomarker_names, collapse = " + "))
)

# Prepare design matrix and outcome for training data for Model B
x_train <- model.matrix(modelB_formula, data = trainData)[, -1]  # remove intercept
y_train <- trainData$recurrence_5yr

# Perform 10-fold cross-validation with cv.glmnet
cv_fit <- cv.glmnet(x_train, y_train, family = "binomial", alpha = 1,
                    nfolds = 10, type.measure = "auc")

# Use the 1-SE rule for lambda selection to achieve a simpler model 
# with performance similar to the minimum-error lambda.
lambda_1se <- cv_fit$lambda.1se

# Display selected lambda
cat("Selected lambda using the 1-SE rule:", lambda_1se, "\n")

# Extract coefficients at lambda_1se
coef_lasso <- coef(cv_fit, s = lambda_1se)
print(coef_lasso)
```

## Variable Selection and Re-fitting Final Model B

```{r}
# Convert coefficients to a tidy data frame and select non-zero coefficients (excluding intercept)
vars_df <- data.frame(
  variable = row.names(coef_lasso),
  coefficient = as.numeric(coef_lasso)
) %>%
  filter(variable != "(Intercept)") %>%
  arrange(desc(abs(coefficient)))

# Only retain variables with non-zero coefficients
vars_selected <- vars_df %>%
  filter(coefficient != 0)

print(vars_selected)

# For re-fitting, ensure clinical covariates are always included.
base_vars <- c("preop_psa", "age", "gleason", "stage_t_n_m")
# Exclude any dummy-coded columns (if present) for the clinical factors from the selection.
bio_vars <- setdiff(vars_selected$variable, base_vars)
bio_vars <- bio_vars[ !grepl("^gleason|^stage_t_n_m", bio_vars) ]

final_var_list <- c(base_vars, bio_vars)
cat("Final variable list for Model B:", paste(final_var_list, collapse = ", "), "\n")

# Re-fit a standard logistic regression using the selected variables on the training set.
final_formula <- as.formula(
  paste("recurrence_5yr ~", paste(final_var_list, collapse = " + "))
)
final_modelB <- glm(final_formula, data = trainData, family = "binomial")
summary(final_modelB)
```

## Build a Summary Table for Model B Coefficients

```{r}
# Create a table of coefficients, Odds Ratios, Confidence Intervals, and p-values
coef_sum <- summary(final_modelB)$coefficients[-1, , drop = FALSE]  # exclude intercept

coef_est <- coef_sum[, 1]
coef_se  <- coef_sum[, 2]
coef_p   <- coef_sum[, 4]

or_est   <- exp(coef_est)
or_lower <- exp(coef_est - 1.96 * coef_se)
or_upper <- exp(coef_est + 1.96 * coef_se)

finalB_table <- data.frame(
  Variable    = row.names(coef_sum),
  Coefficient = round(coef_est, 3),
  StdError    = round(coef_se, 3),
  OddsRatio   = round(or_est, 3),
  CI95        = paste0("(", round(or_lower, 3), ", ", round(or_upper, 3), ")"),
  pValue      = round(coef_p, 4),
  stringsAsFactors = FALSE
)

kable(
  finalB_table,
  caption = "Final Logistic Regression (Model B) with Selected Variables"
)
```


## Model Evaluation on the Test Set

### Predictions and ROC Analysis

```{r}
# Predictions for Model A on test set
predA_test <- predict(modelA_final, newdata = testData, type = "response")
rocA <- roc(testData$recurrence_5yr, predA_test)
aucA <- auc(rocA)
ci_aucA <- ci.auc(rocA, method = "bootstrap", boot.n = 1000)

# Predictions for Model B on test set
predB_test <- predict(final_modelB, newdata = testData, type = "response")
rocB <- roc(testData$recurrence_5yr, predB_test)
aucB <- auc(rocB)
ci_aucB <- ci.auc(rocB, method = "bootstrap", boot.n = 1000)

# Display AUCs with bootstrap 95% confidence intervals
cat("Model A (Clinical Only) AUC:", round(aucA, 3), "with 95% CI", round(ci_aucA[1], 3), "-", round(ci_aucA[3], 3), "\n")
cat("Model B (Clinical + Biomarkers) AUC:", round(aucB, 3), "with 95% CI", round(ci_aucB[1], 3), "-", round(ci_aucB[3], 3), "\n")

# Compare the two models using the DeLong test on the test set
roc_test <- roc.test(rocA, rocB, method = "delong")
print(roc_test)
```

## Plot ROC Curves for Test Set Comparison

```{r}
plot(rocA, col = "blue", main = "ROC Curves Comparison for Model A vs. Model B")
lines(rocB, col = "red")
legend(
  "bottomright",
  legend = c(
    paste0("Model A (Only Covariates): AUC=", round(aucA, 2), ""),
    paste0("Model B (Covariates + Biomarkers): AUC=", round(aucB, 2), "")
  ),
  col = c("blue", "red"),
  lwd = 2
)
```